---
title: "p8105_hw2_dtw2127"
author: Dee Wang 
output: github_document
---

## Problem 1 

```{r setup, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(lubridate)
library(ggplot2)
```

We will read in and clean the "Mr. Trash Wheel" sheet. We'll skip the first row as it has a graphic, clean up the variable names, omit rows with missing data and round the sports balls values to the nearest integer. 

```{r}

mrtrash_data = read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = "Mr. Trash Wheel", skip = 1) %>% #specify sheet to read in and skip row with graphic 
  janitor::clean_names() %>%
  select(dumpster:homes_powered) %>% #omit non-data entries
  drop_na() %>% # omit rows without dumpster-specific data (drop rows with missing data)
  mutate(sports_balls = round(sports_balls)) #round sports balls number to nearest integer
  
```

Next, we'll read and clean precipitation data for 2018 and 2019. 

```{r}
precip_2018 =
  read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = "2018 Precipitation", skip = 1) %>% 
  drop_na() %>% #drop rows with missing data 
  mutate(Year = 2018) #add year variable 
 
precip_2019 =
  read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = "2019 Precipitation", skip = 1) %>% 
  drop_na() %>% #drop rows with missing data 
  mutate(Year = 2019) #add year variable 
 
```

Next we will combine the precipitation datasets and convert month to a character variable. 

```{r}

precip_combined = bind_rows(precip_2018, precip_2019) %>% #combine precipitation datasets
  mutate(Month = month.name[Month]) #convert month to character variable  

```

The Mr. Trash Wheel dataset has `r nrow(mrtrash_data)` observations and `r ncol(mrtrash_data)` variables for the years `r (pull(mrtrash_data, year)) %>% min()` to `r (pull(mrtrash_data, year)) %>% max()`. The dataset includes information on dumpster number, date of collection, weight and volume of litter, as well as how many homes are powered by the trash incinerated, among other variables. 

There are `r nrow(precip_2018)` complete observations in the 2018 Precipitation Dataset, and `r nrow(precip_2019)` complete observations in the 2019 Precipitation Dataset. The combined dataset has `r nrow(precip_combined)` observations.The 2018 Precipitation Dataset has monthly precipitation values (in) for January through December, whereas the 2019 Precipitation Dataset only has monthly precipitation values (in) for January through June. Average monthly rain fall for 2018 was `r pull(precip_2018, Total) %>% mean() %>% round(2)`, and `r pull(precip_2019, Total) %>% mean() %>% round(2)` for the months of January through June in 2019. 

The total precipitation in 2018 was `r pull(precip_2018, Total) %>% sum()`.

The median number of sports balls in a dumpster in 2019 was `r pull(mrtrash_data) %>% median() %>% round(2)`

## Problem 2 

First, we will import and clean the data in pols-month. We will break up the 'mon' variable into 'year', 'month' and 'day' variables, replace month number with month name, create a president variable using the 'gop' and 'dem' variable values and remove the variables 'prez_dem' and 'prez_gop', and remove the 'day' variable. 

```{r}

pols_month = read_csv("./data/pols-month.csv") %>% 
  separate(mon, into = c("Year", "Month", "Day"), sep = "-") %>%
  mutate(Month = as.numeric(Month)) %>% 
  mutate(Month = month.name[Month]) %>% 
  pivot_longer(c(prez_gop,prez_dem), names_prefix = "prez_", names_to = "President") %>% 
  filter(value > 0) %>%
  select(-Day) %>% #remove day variable 
  mutate(Year = as.numeric(Year))

#note -- value of 2 indicates Ford assuming presidency.
```

Next, we will clean the snp dataset similarly to how we cleaned the pols-month dataset. We will replace arrange the dataset according to year and month and will have year and month as the leading columns. 

```{r}

snp <- read_csv("./data/snp.csv") %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  separate(date, into = c("Year", "Month", "Day")) %>% 
  mutate(Year = as.numeric(Year)) %>% 
  mutate(Year = ifelse(Year > 2021, Year - 100, Year)) %>%
  mutate(Month = as.numeric(Month)) %>%
  mutate(Month = month.name[Month]) %>%
  arrange(Year, Month) 

```

Next, we will tidy the unemployment data so that we can merge it with the previous datasets. To ensure that the 'Month' variable is consistent with the pols-month and snp datasets, we will  convert the abbreviated names to a numeric form, and then convert the numeric form to character using month.name. 

```{r}
unemployment = read_csv("./data/unemployment.csv") %>% 
  pivot_longer(
    Jan:Dec,
    names_to = "Month",
    values_to = "Unemployment (%)" 
  ) %>% 
  mutate(Month = match(Month, month.abb)) %>% 
  mutate(Month = month.name[Month])
```

We'll merge snp into pols, and then merge unemployment using left joins. 

```{r}

pols_snp = left_join(pols_month, snp, by = c("Year", "Month"))

pols_snp_unemployment = left_join(pols_snp, unemployment, by = c("Year", "Month"))

```

The original 'pols' dataset consisted of `r nrow(pols_month)` rows, and 9 columns. The dataset has information related to the number of democratic and republican national politicians at a given time. 

The original 'snp' dataset contained `r nrow(snp)` rows and 2 columns. The dataset contains information about closing values of the S&P stock index at a given time.

The original 'unemployment' dataset contained `r nrow(unemployment)` and 13 columns. The dataset contains information about percentage of unemployment at a given time. 

The combined dataset is `r nrow(pols_snp_unemployment)` rows and `r ncol(pols_snp_unemployment)`, and contains data from the years `r pull(pols_snp_unemployment, Year) %>% min()` to `r pull(pols_snp_unemployment, Year) %>% max()`. Key variables include month and year, closing value of the SNP, party affiliation of the president and unemployment percentage.

## Problem 3 

We will load and tidy the 'popular baby names' dataset. The Ethnicity categories are not consistently used. We need to change instances of "ASIAN AND PACI" to "ASIAN AND PACIFIC ISLANDER", and "BLACK NON HISP" to "BLACK NON HISPANIC", and "WHITE NON HISP" to "WHITE NON HISPANIC". We also want the case of the names to be consistent. 

```{r}
babynames <- read_csv("./data/Popular_Baby_Names.csv") %>% 
  mutate(Ethnicity = replace(Ethnicity, Ethnicity == "BLACK NON HISP", "BLACK NON HISPANIC")
  ) %>% 
  mutate(Ethnicity = replace(Ethnicity, Ethnicity == "WHITE NON HISP", "WHITE NON HISPANIC")
  ) %>% 
  mutate(Ethnicity = replace(Ethnicity, Ethnicity == "ASIAN AND PACI", "ASIAN AND PACIFIC ISLANDER")
  ) %>% 
  mutate("Child's First Name" = str_to_title(`Child's First Name`)) %>%
  unique() #remove rows that are duplicated
  
#this could probably be done more efficiently
```

We'll create reader-friendly tables showing the rank in popularity of the name "Olivia" as a female baby name over time. 

```{r}
popularity_olivia <- babynames %>% 
  filter(`Child's First Name` == "Olivia") %>% 
  select("Ethnicity", "Year of Birth", "Rank") %>% 
  pivot_wider(names_from = "Year of Birth",
            values_from = "Rank") 

popularity_olivia = popularity_olivia[, c(1,5,4,3,2)] #reorder columns

```

Now we will create a table showing the most popular male baby name over time. 

```{r}
popular_male_babyname <- babynames %>% 
  filter(Gender == "MALE") %>% 
  select("Ethnicity", "Year of Birth", "Rank", "Child's First Name") %>% 
  pivot_wider(names_from = "Year of Birth",
              values_from = `Child's First Name`) %>% 
  filter(Rank == 1) %>% 
  select(-Rank)

```

Next we'll create a scatter plot showing number of male, white, non-hispanic children born in 2016 with a certain name against the rank in popularity of that name. 

```{r}

white_male_names_2016 = babynames %>%
  filter(Gender == "MALE" & Ethnicity == "WHITE NON HISPANIC" & `Year of Birth` == 2016) %>% 
  select(Count, Rank)

ggplot(white_male_names_2016, aes(x = Count, y = Rank)) + 
  geom_point() + 
  labs( 
    title = "Frequency of white non-hispanic male baby names against rank in 2016")

```

