---
title: "p8105_hw2_dtw2127"
author: Dee Wang 
output: github_document
---

## Problem 1 

We will read in and clean the "Mr. Trash Wheel" sheet.

```{r}
library(readxl)
library(tidyverse)
library(lubridate)

mrtrash_data = read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = "Mr. Trash Wheel", skip = 1) %>% #specify sheet to read in and skip row with graphic 
  select("Dumpster":"Homes Powered*") %>% #omit non-data entries
  drop_na() %>% # omit rows without dumpster-specific data (drop rows with missing data)
  mutate(`Sports Balls` = round(`Sports Balls`)) #round sports balls number to nearest integer
  
```

Next, we'll read and clean precipitation data for 2018 and 2019. 

```{r}
precip_2018 =
  read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = "2018 Precipitation", skip = 1) %>% 
  drop_na() %>% #drop rows with missing data 
  mutate(Year = 2018) #add year variable 
 
precip_2019 =
  read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = "2019 Precipitation", skip = 1) %>% 
  drop_na() %>% #drop rows with missing data 
  mutate(Year = 2019) #add year variable 
 
```

Next we will combine the precipitation datasets and convert month to a character variable. 

```{r}

precip_combined = bind_rows(precip_2018, precip_2019) %>% #combine precipitation datasets
  mutate(Month = month.name[Month]) #convert month to character variable  

```

There are `r nrow(precip_2018)` complete observations in the 2018 Precipitation Dataset, and `r nrow(precip_2019)` complete observations in the 2019 Precipitation Dataset. The combined dataset has `r nrow(precip_combined)` observations.

The 2018 Precipitation Dataset has monthly precipitation values (in) for January through December, whereas the 2019 Precipitation Dataset only has monthly precipitation values (in) for January through June. Average monthly rain fall for 2018 was `r pull(precip_2018, Total) %>% mean() %>% round(2)`, and `r pull(precip_2019, Total) %>% mean() %>% round(2)` for the months of January through June in 2019. The total precipitation in 2018 was `r pull(precip_2018, Total) %>% sum()`.

The median number of sports balls in a dumpster in 2019 was `r pull(mrtrash_data) %>% median() %>% round(2)`

## Problem 2 

First, we will import and clean the data in pols-month.csv.

```{r}

pols_month = read_csv("./data/pols-month.csv") %>% 
  separate(mon, into = c("Year", "Month", "Day"), sep = "-") %>%
  mutate(Month = as.numeric(Month)) %>% 
  mutate(Month = month.name[Month]) %>% 
  pivot_longer(c(prez_gop,prez_dem), names_prefix = "prez_", names_to = "President") %>% 
  filter(value > 0) %>%
  select(-Day) #remove day variable 

#note -- value of 2 indicates Ford assuming presidency.
```

```{r}

snp <- read_csv("./data/snp.csv") %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  separate(date, into = c("Year", "Month", "Day")) %>% 
  mutate(Month = as.numeric(Month)) %>%
  mutate(Month = month.name[Month]) %>%
  arrange(Year, Month) 

## need to figure out how to fix the dates (>= 2500)

```

Next, we will tidy the unemployment data. 

```{r}
unemployment = read_csv("./data/unemployment.csv") %>% 
  pivot_longer(
    Jan:Dec,
    names_to = "Month",
    values_to = "Unemployment (%)" 
  ) %>% 
  mutate(Month = match(Month, month.abb)) %>% 
  mutate(Month = month.name[Month])
```

## Problem 3 

We will load and tidy the 'popular baby names' dataset. The Ethnicity categories are not consistently used. We need to change instances of "ASIAN AND PACI" to "ASIAN AND PACIFIC ISLANDER", and "BLACK NON HISP" to "BLACK NON HISPANIC", and "WHITE NON HISP" to "WHITE NON HISPANIC". 

```{r}
babynames <- read_csv("./data/Popular_Baby_Names.csv") %>% 
  mutate(Ethnicity = replace(Ethnicity, Ethnicity == "BLACK NON HISP", "BLACK NON HISPANIC")
  ) %>% 
  mutate(Ethnicity = replace(Ethnicity, Ethnicity == "WHITE NON HISP", "WHITE NON HISPANIC")
  ) %>% 
  mutate(Ethnicity = replace(Ethnicity, Ethnicity == "ASIAN AND PACI", "ASIAN AND PACIFIC ISLANDER")
  ) %>% 
  unique() #remove rows that are duplicated
  
#this could probably be done more efficiently
```

We'll create reader-friendly tables showing the rank in popularity of the name "Olivia" as a female baby name over time. 

```{r}
popularity_olivia <- babynames %>% 
  filter(`Child's First Name` == "Olivia") %>% 
  select("Ethnicity", "Year of Birth", "Rank") %>% 
  pivot_wider(names_from = "Year of Birth",
            values_from = "Rank") 

popularity_olivia = popularity_olivia[, c(1,5,4,3,2)] #reorder columns

```

