---
title: "p8105_hw2_dtw2127"
author: Dee Wang 
output: github_document
---

## Problem 1 

We will read in and clean the "Mr. Trash Wheel" sheet.

```{r}
library(readxl)
library(tidyverse)

mrtrash_data = read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = "Mr. Trash Wheel", skip = 1) %>% #specify sheet to read in and skip row with graphic 
  select("Dumpster":"Homes Powered*") %>% #omit non-data entries
  drop_na() %>% # omit rows without dumpster-specific data (drop rows with missing data)
  mutate(`Sports Balls` = round(`Sports Balls`)) #round sports balls number to nearest integer
  
```

Next, we'll read and clean precipitation data for 2018 and 2019. 

```{r}
precip_2018 =
  read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = "2018 Precipitation", skip = 1) %>% 
  drop_na() %>% #drop rows with missing data 
  mutate(Year = 2018) #add year variable 
 
precip_2019 =
  read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", sheet = "2019 Precipitation", skip = 1) %>% 
  drop_na() %>% #drop rows with missing data 
  mutate(Year = 2019) #add year variable 
 
```

Next we will combine the precipitation datasets and convert month to a character variable. 

```{r}

precip_combined = bind_rows(precip_2018, precip_2019) %>% #combine precipitation datasets
  mutate(Month = month.name[Month]) #convert month to character variable  

```

There are `r nrow(precip_2018)` complete observations in the 2018 Precipitation Dataset, and `r nrow(precip_2019)` complete observations in the 2019 Precipitation Dataset. The combined dataset has `r nrow(precip_combined)` observations.

The 2018 Precipitation Dataset has monthly precipitation values (in) for January through December, whereas the 2019 Precipitation Dataset only has monthly precipitation values (in) for January through June. Average monthly rain fall for 2018 was `r pull(precip_2018, Total) %>% mean() %>% round(2)`, and `r pull(precip_2019, Total) %>% mean() %>% round(2)` for the months of January through June in 2019. The total precipitation in 2018 was `r pull(precip_2018, Total) %>% sum()`.

The median number of sports balls in a dumpster in 2019 was `r pull(mrtrash_data) %>% median() %>% round(2)`


